What we {\sf eval} in the shadows

### 2021-03-31 paper/main.tex
related

\subsection{R in a nutshell}

R is a lazy functional programming language~\cite{ecoop12} with
dynamic features that allow to write object-oriented code. Most data
types are vectorized. Values are constructed by the \c{c(...)}
function, e.g. \c{c("a","bc")} creates a vector of two strings. To
enable equational reasoning, R copies values accessible through
multiple variables when they are written to. Values can be tagged by
user-defined attributes. For instance, one can attach the attribute
\c{dim} to the value \c{x<-c(1,2,3,4)} with \c{attr(x,"dim")<-c(2,2)}.
This causes arithmetic functions to treat \c x as a $2 \times 2$
matrix. Another attribute is \c{class} which can be bound to a list of
names, e.g., \c{class(x)<-"human"}. This sets the class of \c{x} to
\c{human}; classes are used for object-oriented dispatch. Every R
linguistic construct is desugared to a function call, even control
flow statements, assignments, and bracketing. Furthermore, all
functions can be redefined in user code. This makes R both flexible
and challenging to compile~\cite{dls19}. Arguments to user-defined
functions are bundled into thunks called \emph{promises}. Logically, a
promise combines an expression's code, its environment, and its value.
To access the value of a promise, one must force it. Forcing a promise
triggers evaluation and the computed value is captured for future
reference.

\subsection{Eval Semantics}

The expressive power of \eval depends on design decisions along two
axes:
\begin{itemize}
\item {\bf Scoping:} What environment does \eval executes its argument
  in? JavaScript and R evaluate it in the current environment, thus
  exposing local variables and parameters and breaking the caller's
  abstraction boundary. Julia is more restrictive, \eval runs at the
  ``top level'', in the global environment. JavaScript has a global
  \eval that behaves like Julia, and a ``strict mode'' in which \eval
  may access local variables, but fobidden from injecting new ones in
  the local environment. In Java, one can implement \eval with
  Julia-semantics.\footnote{While \eval used to be the purview of
  interpreted languages, just-in-time compilation lifted this
  restriction. \Eval takes its input, wraps it in a static method of a
  new anonymous class, generates bytecode for that class, invokes
  the class loader to install that code, and finally reflectively
  calls the method. }
\item {\bf Reflective API:} What other reflective operations are
  exposed to user code? JavaScript provides few reflective functions
  (other than enumeration of an object's properties and string-based
  indexing), the impact of \eval is thus limited. R, on the other
  hand, has a rich reflective API. For example, R allows user code to
  walk the call stack and arbitrarily add and delete local variables.
  This means that there are few limits to the side-effects of an
  \eval.
\end{itemize}

>>>>

\subsection{R, Briefly}

\citet{ecoop12} gave a programming language-centric overview of the R language.
They characterized it as a lazy, vectorized, functional language with rich
dynamic features that allow layering several object systems. Most data types are
sequences of primitive values constructed by calling the combination function
\c{c}, so that \c{c(1,2)} evaluates to a vector of two strings, and constants
such as \c{42} are vectors of length one. To enable equational reasoning, values
accessible through multiple aliases are copied when written to. Furthermore,
values can be tagged by user-defined attributes, key-value pairs. For instance,
the attribute \c{dim}-\c{c(2,2)} can be attached to the value at \c{x} by the
call \c{attr(x,"dim")<-c(2,2)}. In this case, the addition of this attribute
ensure that the vector \c x will be treated as a $2 \times 2$ matrix by all
arithmetic operations. The \c{class} attribute is used to give a 'class', in the
object-oriented sense, to values. So, \c{class(x)<-"human"} sets the class of
\c{x} to \c{human}; classes are used for method dispatch. Every linguistic
construct is desugared to a function call, even control flow statements,
assignments, and bracketing. All functions can be shadowed or redefined, making
the language at the same time remarkably flexible and exceeding challenging to
compile statically as vividly detailed by~\citet{dls19}. R uses a relaxed call-by-need convention for
passing arguments to functions. Each argument is a thunk composed of an
expression, its environment, and a slot for the result, these are called
\emph{promises}. To get the value of an argument, the corresponding promise must
be forced, once forced the promise's result is cached for future use.

\subsection{The Semantics of Eval and Friends}

While a facility to transform data into code is available in many languages,
some design choices affect the expressive power of \eval and related functions.
The key design choices that we review here are the input format, the environment
in which \eval-generated code is evaluated and the reflective operations that
are available to that code. Fig.~\ref{comp} summarizes design choices for
several popular languages.

\begin{figure}[!t]\center\small
\begin{tabular}{r|l|l|l}\hline
\tiny\sc Language&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\hline
\bf Julia     & expression& toplevel         & ??\\
\bf Java      & bytecode  & classloader       & data\\
\bf JavaScript& text      & current, toplevel& data\\
\bf R         & expression& programmatic      & data, stack, envir.\\\hline
\end{tabular}
\caption{Design space of \eval}\label{comp}
\end{figure}

The input to \eval can be any format that can be turned into code. Both Julia
and R mandate that the input format be parsed expressions. JavaScript allows any
expression as a character string. Java is the most restrictive as a classloader
only accepts complete classes in the bytecode format recognized by the virtual
machine.

The environment in which \eval executes the code can be the lexically enclosing
environment of \eval (current), the global environment (toplevel), a limited
global environment (classloader), or an environment chosen by the caller of
\eval (programmatic). The most restrictive variant is that of Java, where the
scope of newly loaded code is the set of classes exposed by the current class
loader. Julia is the next most restrictive, as only global definitions can be
observed and modified by \eval. JavaScript has a variant of \eval that runs in
the current scope and a most strict mode that is restricted to global variables.
Finally, R allows the caller to choose which environment to use with \eval.

WORK IN  PROGREESS

JavaScript and R evaluate it in the current environment, thus exposing local
variables and parameters and breaking the caller's abstraction boundary. Julia
is more restrictive, \eval runs at the ``top level'', in the global environment.
JavaScript has a global \eval that behaves like Julia, and a ``strict mode'' in
which \eval may access local variables, but fobidden from injecting new ones in
the local environment. In Java, one can implement \eval with
Julia-semantics.\footnote{While \eval used to be the purview of interpreted
languages, just-in-time compilation lifted this restriction. \Eval takes its
input, wraps it in a static method of a new anonymous class, generates bytecode
for that class, invokes the class loader to install that code, and finally
reflectively calls the method. }

 {\bf Reflective API:} What other reflective operations are
  exposed to user code? JavaScript provides few reflective functions
  (other than enumeration of an object's properties and string-based
  indexing), the impact of \eval is thus limited. R, on the other
  hand, has a rich reflective API. For example, R allows user code to
  walk the call stack and arbitrarily add and delete local variables.
  This means that there are few limits to the side-effects of an
  \eval.
